The Unicode Cookbook \linebreak for Linguists

### 2018-06-05 book/chapters/unicode_approach.tex
fixes abbreviations, address paperhivers through chapter 1

Each version of the Unicode Standard (\citealp{Unicode2017}, as of writing at
version 10) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Different from
many other documents of the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthography traditions all over the world.} The \textsc{core
specification} is a book directed toward human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
humanly readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, Web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

@@

The Unicode Standard is a \textsc{character encoding system} which
goal it is to support the interchange and processing of written characters and
text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems. The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e. the code point, which can be encoded differently by different encoding systems.} Underlyingly, the character encoding is
represented by a range of numerical values called a \textsc{code space}, which
is used to encode a set of characters. A \textsc{code point} is a unique
non-negative integer within a code space (i.e.~within a certain numerical
range). In the Unicode Standard character encoding system, an \textsc{abstract
character}, for example the \textsc{latin small letter p}, is mapped to a
particular code point, in this case the decimal value 112, normally represented in
hexadecimal, which then looks in Unicode parlance as
\uni{0070}.
%\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
%(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
%(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
%computers may treat this code point as an 8-bit binary (base-2) sequence (11100000), as
%can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
%(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
%16 = 112$.} 
That encoded abstract character is rendered on a computer screen (or
printed page) as a \textsc{glyph}, e.g. <p>, depending on the \textsc{font} and
the context in which that character appears.

@@

The Unicode Standard provides technological specifications for creating locale specific data
in so-called \textsc{Unicode Locale Descriptions}, i.e.~a set of specification
that defines a set of language-specific elements (e.g.~tailored grapheme
clusters, collation order, capitalization-equivalence), as well as other special
information, like how to format numbers, dates, or currencies. Locale
descriptions are saved in the \textsc{Common Locale Data Repository
(CLDR)},\footnote{More information about the CLDR can be found here:
\url{http://cldr.unicode.org/}.} a repository of
language-specific definitions of writing system properties, each of which
describes specific usages of characters. Each locale can be encoded in a
document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
format and vocabulary for the exchange of structured locale data. Unicode Locale
Descriptions allow users to define language- or even resource-specific writing
systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
system} only very loosely, as it is not a central concept in the Unicode
Standard. A writing system is, ``A set of rules for using one or more scripts to
write a particular language. Examples include the American English writing
system, the British English writing system, the French writing system, and the
Japanese writing system.''} However, there are various drawbacks of locale
descriptions for the daily practice of linguistic work in a multilingual setting.

>>>>

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Different from
many other documents of the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthography traditions all over the world.} The \textsc{core
specification} is a book directed toward human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
humanly readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, Web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

@@

The Unicode Standard is a \textsc{character encoding system} which
goal it is to support the interchange and processing of written characters and
text in a computational setting.\footnote{An insightful reviewer notes that the term \textsc{encoding} is used for both sequences of code points and text encoded as bit patterns. Hence a Unicode-aware programmer might prefer to say that UTF-8, UTF-16, etc., are Unicode encoding systems. The issue is that the Unicode Standard introduces a layer of indirection between characters and bit patterns, i.e.\ the code point, which can be encoded differently by different encoding systems.} Underlyingly, the character encoding is
represented by a range of numerical values called a \textsc{code space}, which
is used to encode a set of characters. A \textsc{code point} is a unique
non-negative integer within a code space (i.e.~within a certain numerical
range). In the Unicode Standard character encoding system, an \textsc{abstract
character}, for example the \textsc{latin small letter p}, is mapped to a
particular code point, in this case the decimal value 112, normally represented in
hexadecimal, which then looks in Unicode parlance as
\uni{0070}.
%\footnote{Hexadecimal (base-16) 0070 is equivalent to decimal
%(base-10) 112, which can be calculated by considering that $(0\cdot16^3) +
%(0\cdot16^2) + (7\cdot16^1) + (0\cdot16^0) = 7\cdot16 = 112$. Underlyingly,
%computers may treat this code point as an 8-bit binary (base-2) sequence (11100000), as
%can be seen by calculating that $(1\cdot2^7) + (1\cdot2^6) + (1\cdot2^5) +
%(0\cdot2^4) + (0\cdot2^3) + (0\cdot2^2) + (0\cdot2^1) + (0\cdot2^0) = 64 + 32 +
%16 = 112$.} 
That encoded abstract character is rendered on a computer screen (or
printed page) as a \textsc{glyph}, e.g.\ <p>, depending on the \textsc{font} and
the context in which that character appears.

@@

The Unicode Standard provides technological specifications for creating locale specific data
in so-called \textsc{Unicode Locales}, i.e.~a set of specification
that defines a set of language-specific elements (e.g.~tailored grapheme
clusters, collation order, capitalization-equivalence), as well as other special
information, like how to format numbers, dates, or currencies. Locale
descriptions are saved in the \textsc{Common Locale Data Repository
(CLDR)},\footnote{More information about the CLDR can be found here:
\url{http://cldr.unicode.org/}.} a repository of
language-specific definitions of writing system properties, each of which
describes specific usages of characters. Each locale can be encoded in a
document using the \textsc{Locale Data Markup Language (LDML)}. LDML is an XML
format and vocabulary for the exchange of structured locale data. Unicode Locale
Descriptions allow users to define language- or even resource-specific writing
systems or orthographies.\footnote{The Unicode Glossary defines \textsc{writing
system} only very loosely, as it is not a central concept in the Unicode
Standard. A writing system is, ``A set of rules for using one or more scripts to
write a particular language. Examples include the American English writing
system, the British English writing system, the French writing system, and the
Japanese writing system.''} However, there are various drawbacks of locale
descriptions for the daily practice of linguistic work in a multilingual setting.

### 2018-06-05 book/chapters/use_cases.tex
comments out some stuff we probably won't use in the use cases
fixes abbreviations, address paperhivers through chapter 1

\section{Cross-orthographic analysis of writing systems}

@@

\begin{comment}

@@

The most basic overall Unicode character property is the \textsc{General
Category}, which categorizes Unicode characters into: \emph{Letters,
Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*. Each Unicode
character property also has a character property value. So each General Category
is denoted by a property value by a single letter abbreviation, e.g. ``L'' for
Letter. Each General Category property also has subcategories that are also
properties with property values. For example, ``Letters'' is broken down into
Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier
Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter''
includes characters such as uncased letters as in the Chinese writing system.}
By defining character properties for each code point, algorithmic
implementations such as text processing applications or regular expression
engines that are conformant to the Unicode Standard can be used to recognize
whole (sub)categories of characters or ranges of characters, e.g.~to match
proper names in English in title case or more generally to match character
types, such as scripts, or symbols and integers in email addresses, phone
numbers, etc.

Script information is useful for algorithmic processes, such as collation,
searching and for multilingual text processing. For example, if a regular
expression engine includes functionality to match characters at the level of
script, then it provides the ability to identify characters, e.g. \match{Greek},
which tests whether letters in a text belong to the Greek script or not. This
functionality can be used to determine the boundaries between different writing
systems that appear in the same document. In Section 6 (Use cases of grapheme
tokenization) we provide an example of a regular expression match at the level
of character property.

@@

\section{PHOIBLE}

@@

\citet{Moran2012;MoranMcCloy2014} mark tones as singletons with Space Modifier
Letters, e.g. <˦> for a phonemic high tone, instead of accent diacritics,
alleviating potential conflicts.

>>>>

\begin{comment}
\section{Cross-orthographic analysis of writing systems}

@@

@@

The most basic overall Unicode character property is the \textsc{General
Category}, which categorizes Unicode characters into: \emph{Letters,
Punctuation, Symbols, Marks, Numbers, Separators, }and* Other*. Each Unicode
character property also has a character property value. So each General Category
is denoted by a property value by a single letter abbreviation, e.g.\ ``L'' for
Letter. Each General Category property also has subcategories that are also
properties with property values. For example, ``Letters'' is broken down into
Uppercase Letter (Lu), Lowercase Letter (Ll), Titlecase Letter (Lt), Modifier
Letter (Lm) and Other Letter (Lo).\footnote{The subcategory ``Other Letter''
includes characters such as uncased letters as in the Chinese writing system.}
By defining character properties for each code point, algorithmic
implementations such as text processing applications or regular expression
engines that are conformant to the Unicode Standard can be used to recognize
whole (sub)categories of characters or ranges of characters, e.g.~to match
proper names in English in title case or more generally to match character
types, such as scripts, or symbols and integers in email addresses, phone
numbers, etc.

Script information is useful for algorithmic processes, such as collation,
searching and for multilingual text processing. For example, if a regular
expression engine includes functionality to match characters at the level of
script, then it provides the ability to identify characters, e.g.\ \match{Greek},
which tests whether letters in a text belong to the Greek script or not. This
functionality can be used to determine the boundaries between different writing
systems that appear in the same document. In Section 6 (Use cases of grapheme
tokenization) we provide an example of a regular expression match at the level
of character property.

@@

\begin{comment}
\section{PHOIBLE}

@@

\citet{Moran2012;MoranMcCloy2014} mark tones as singletons with Space Modifier
Letters, e.g.\ <˦> for a phonemic high tone, instead of accent diacritics,
alleviating potential conflicts.
